---
title: "American Airlines departure delay study case"
author: "Javier Cano Montero"
date: "26 de febrero de 2017"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Objective

In the present study case we will try to understand the reasons why departure delays occur for American Airlines in airports which operate 5000 or more of its flights and will focus on its worst performance airport (San Antonio).

# Summary

* Number of flights an airport is operating is not correlated with departure delays.
* A late aircraft is highly correlated with departure delays.
* A late aircraft is highly correlated with taxiing times.
* A late aircraft is highly correlated with short flights.
* American Airlines in San Antonio operates mostly short flights, making its planes departure times to be among the most delayed.
* American Airlines in San Antonio is one of the carriers that spends more time taxiing, making its planes to arrive late.

# Setting up the data

In this problem set we will use the data for the flights in 2016 that the Bureau of Transportation Statistics of the American Department of Transportation published on <https://www.transtats.bts.gov/Tables.asp?DB_ID=120&DB_Name=Airline%20On-Time%20Performance%20Data&DB_Short_Name=On-Time>.

First we will load the libraries and read the data. Because the data we are using is really heavy, two auxiliar functions have been provided to automatically download an unzip the data.

```{r setup, message=FALSE}
# We load the libraries and the data required to run the script and generate this document.
rm(list=ls())
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
source("aux_functions.R")

# Uncomment to download and unzip data into data folder (may take a while)
# getData()
# Uncomment to merge monthly files into a yearly file (may take a while)
# mergeFiles()

perf <- fread("../data/On_Time_On_Time_Performance_2016_full.csv", header = T, data.table = FALSE, stringsAsFactors = T, showProgress = FALSE)
```

After reading the data, we will split it so we can focus on American Airlines and make comparisions with the other carriers performances.

```{r splitting the data,  message=FALSE}
# Now we split the data so we can focus on American Airlines
# Data from AA
perf_aa <- filter(perf,Carrier=="AA")
perf_aa <- droplevels(perf_aa)
# Data from another carriers
perf <- filter(perf,Carrier!="AA")
perf <- droplevels(perf)
```

# Finding the airport with the most delay mean time

To understand why the departure delays happen, we will study big airports with 5000 or more departures during 2016. To achieve this goal we will calculate the mean of the following features:

* DepDelayMinutes: Difference in minutes between scheduled and actual departure time. Early departures set to 0.
* CarrierDelay: Carrier Delay, in Minutes.
* WeatherDelay: Weather Delay, in Minutes
* NASDelay: National Air System Delay, in Minutes.
* SecurityDelay: Security Delay, in Minutes.
* LateAircraftDelay: Late Aircraft Delay, in Minutes.

```{r first analysis,  message=FALSE}
# Departures by airport
# AA
departures_aa <- 
  perf_aa %>% 
  group_by(OriginCityName) %>% 
  count() 

departures_aa <- 
  departures_aa %>% 
  filter(n>5000)
# Others
departures_others <- 
  perf %>% 
  filter(OriginCityName %in% departures_aa$OriginCityName) %>% 
  group_by(OriginCityName) %>% 
  count() 

# Informed departure delay
# AA
mean_delay_aa <- 
  perf_aa %>% 
  filter(!is.na(DepDelayMinutes) &
           DepDelayMinutes>0 &
           !is.na(CarrierDelay) &
           !is.na(WeatherDelay) &
           !is.na(NASDelay) &
           !is.na(SecurityDelay) &
           !is.na(LateAircraftDelay)
  ) %>%
  group_by(OriginCityName) %>% 
  select(OriginCityName,DepDelayMinutes) %>%
  summarise_each(funs=c("mean","sd")) 
# Others
mean_delay_others <- 
  perf %>% 
  filter(OriginCityName %in% departures_aa$OriginCityName &
           !is.na(DepDelayMinutes) &
           DepDelayMinutes>0 &
           !is.na(CarrierDelay) &
           !is.na(WeatherDelay) &
           !is.na(NASDelay) &
           !is.na(SecurityDelay) &
           !is.na(LateAircraftDelay)
  ) %>%
  group_by(OriginCityName) %>% 
  select(OriginCityName,DepDelayMinutes) %>%
  summarise_each(funs=c("mean","sd")) 

# Why is AA having delays?
delays_by_airport_aa <- 
  perf_aa %>% 
  filter(!is.na(DepDelayMinutes) & 
           DepDelayMinutes>0 &
           !is.na(CarrierDelay) &
           !is.na(WeatherDelay) &
           !is.na(NASDelay) &
           !is.na(SecurityDelay) &
           !is.na(LateAircraftDelay)
  ) %>%  
  group_by(OriginCityName) %>% 
  select(OriginCityName,DepDelayMinutes,CarrierDelay,WeatherDelay,NASDelay,SecurityDelay,LateAircraftDelay) %>% 
  summarise_each(funs=c("mean"))

# Why are the other carriers having delays?
delays_by_airport_others <- 
  perf %>% 
  filter(OriginCityName %in% departures_aa$OriginCityName &
           !is.na(DepDelayMinutes) & 
           DepDelayMinutes>0 &
           !is.na(CarrierDelay) &
           !is.na(WeatherDelay) &
           !is.na(NASDelay) &
           !is.na(SecurityDelay) &
           !is.na(LateAircraftDelay)
  ) %>%  
  group_by(OriginCityName) %>% 
  select(OriginCityName,DepDelayMinutes,CarrierDelay,WeatherDelay,NASDelay,SecurityDelay,LateAircraftDelay) %>% 
  summarise_each(funs=c("mean"))
```

Now that we have the KPI set we will arrange the airports so we can check the worst 10 with the most mean delayed departure minutes.

```{r most delayed airports, message=FALSE, }
# Let's join the number of departures with the reason of the delays
# AA
delays_by_airport_aa <- 
  departures_aa %>% 
  left_join(delays_by_airport_aa) %>% 
  arrange(desc(DepDelayMinutes))

delays_by_airport_aa[is.na(delays_by_airport_aa)] <- 0

# Others
delays_by_airport_others <- 
  departures_others %>% 
  left_join(delays_by_airport_others)%>% 
  arrange(desc(DepDelayMinutes))

delays_by_airport_others[is.na(delays_by_airport_others)] <- 0

print(delays_by_airport_aa)
```

We can see that the worst airport for American Airlines is San Antonio in Texas with almost 83 minutes of delays to departure flights. 

# Why is this happening?

To find the reason of these delays we will look for a how the features we set correlate among them. We suspect that the number of flights may have an influence on the delays. More number of flights may make the airport to run slower.

Let's find out. First with the data of American Airlines.

```{r finding a reason AA cor n, message=FALSE, }
cor(delays_by_airport_aa[,2:3])[1,]
```

Now with the other carriers' data.

```{r finding a reason others cor n, message=FALSE, }
cor(delays_by_airport_others[,2:3])[1,]
```

We were wrong on our first try. As we can see on the previous correlations, the number of flights is negative correlated with de departure delay. This means that the more flights depart from an airport it suffers less departure delays. This is true also to the other carriers.

We will use the others features to find out the reasons after a departure delay. To do this we will run a linear regression with these features and the departure delay.
Again, first with the data of American Airlines.

```{r finding a reason AA cor ddm, message=FALSE, }
x <- delays_by_airport_aa[,2:8]
summary(glm(data = x, DepDelayMinutes~.))
```

As we can see, the delays induced by the carrier, the weather, tha NAS and a late aricraft are very significant to explain a departure delay. Since we can do nothing to change the weather or the reasons why the NAS delays a flight, we will work with the carrier delays and late aircrafts.


We will try and check if the distance of the flight is also significant to the delay, since long flights may take longer to get ready and take more time flying. To find out if this assumption is true, we will run some more regressions.

```{r checking if we a right 1, message=FALSE }
delay <- summarise(perf_aa %>% group_by(DistanceGroup),
                   count = n(),
                   dist = mean(Distance, na.rm = TRUE),
                   delay = mean(DepDelayMinutes, na.rm=TRUE),
                   carrierDelay = mean(CarrierDelay, na.rm = TRUE),
                   lateCraft = mean(LateAircraftDelay,na.rm=TRUE)
                   )

ggplot(delay, aes(dist, delay)) +
  geom_point(aes(size = count), alpha = 1/2) +
  labs(x="Distance (miles)", y="Delay (mins.)") +
  geom_smooth(method = 'glm') +
  scale_size_area() +
  ggtitle("Distance and delay relation") +
  scale_radius(name="N. Flights")

```

In the previous plot, we can see that the distance is slightly positively correlated with a departure delay. Longer flights have a bit more departure delay.

```{r checking if we are right 2, message=FALSE}
ggplot(delay, aes(dist, carrierDelay)) +
  geom_point(aes(size = count), alpha = 1/2) +
  labs(x="Distance (miles)", y="Carrier delay (mins.)") +
  geom_smooth(method = 'glm') +
  scale_size_area() +
  ggtitle("Distance and carrier delay relation") +
  scale_radius(name="N. Flights") 
```

In this plot, we can see that the delays caused by the carrier and the distance is again positively correlated. Longer flights are more delayed because of the carrier than shorter flights.

```{r checking if we are right 3, message=FALSE}
ggplot(delay, aes(dist, lateCraft)) +
  geom_point(aes(size = count), alpha = 1/2) +
  labs(x="Distance (miles)", y="Late aircraft (mins.)") +
  geom_smooth(method = 'glm') +
  scale_size_area() +
  ggtitle("Distance and late aircraft relation") +
  scale_radius(name="N. Flights") 

```

In this final plot, we can see that a late aircraft delay is more common in shorter flights than in long flights. Why? Checking the departure/arrival performance features in the dataset we find TaxiOut and TaxiIn. These features are the minutes the aircraft takes to taxi from the terminal to the runway (TaxiOut) and the opposite (TaxiIn). Let's check if the time of taxiing is significant with the delays caused by a late aircraft.

```{r short flights delay, message=FALSE}
taxi <- data_frame(perf_aa$LateAircraftDelay,perf_aa$TaxiOut+perf_aa$TaxiIn)
taxi <- taxi[complete.cases(taxi),]
colnames(taxi) <- c("delay","taxi")
summary(glm(data=taxi,delay~.))
```

As we can see, this feature is very significant to explain a late aircraft delay. In the previous plot we saw that shorter flights are more common, and the same plane has to take off and land several times a day, so taxiing happens more often, acummulating little delays.

# San Antonio

As we saw, San Antonio is the airport with more departure delays for American Airlines. Let's see how other carriers behave in this same airport.
First, let's get ready the data.

```{r data for San Antonio, message=FALSE, }
ps_by_carrier <- rbind(perf,perf_aa) %>%
  filter(!is.na(CarrierDelay) &
           OriginCityName=="San Antonio, TX"
  ) %>%
  group_by( Carrier)
```

Now we will check how American Airlines and other carriers operate in San Antonio Airport.

## Flight distances

```{r plotting San Antonio Distance, message=FALSE}
pd <- ggplot(ps_by_carrier, aes(Carrier, Distance)) +
  geom_boxplot(outlier.colour = "red") +
  labs(x="Carrier",y="Distance") +
  ggtitle("Flight distance in San Antonio, TX")

ylim1 = boxplot.stats(ps_by_carrier$Distance)$stats[c(1, 5)]

pd <- pd + coord_cartesian(ylim = ylim1*1.5)

plot(pd)
```

American Airlines departure flights from San Antonio Airport are usually short flights.

## Departure delay

```{r plotting San Antonio DepDelay, message=FALSE}
p0 <- ggplot(ps_by_carrier, aes(Carrier, DepDelayMinutes)) +
  geom_boxplot(outlier.shape=NA) +
  labs(x="Carrier",y="Delay") +
  ggtitle("Delay in San Antonio, TX")

ylim1 = boxplot.stats(ps_by_carrier$DepDelayMinutes)$stats[c(1, 5)]

p0 <- p0 + coord_cartesian(ylim = ylim1*1.5)

plot(p0)
```

We can see that American Airlines is one of the carriers with more departure delays (about 50 minutes in median). American Airlines is one of the carriers with most departure delays in this airport. As we saw previously, short flights have more departure delays than long flights. As American Airlines' flights are mostly short this result is consistent with the previous. 

## Carrier delay

```{r plotting San Antonio CarrierDelay, message=FALSE}
p1 <- ggplot(ps_by_carrier, aes(Carrier, CarrierDelay)) +
  geom_boxplot(outlier.shape=NA) +
  labs(x="Carrier",y="Delay") +
  ggtitle("Delay caused by carrier in San Antonio, TX")

ylim1 = boxplot.stats(ps_by_carrier$CarrierDelay)$stats[c(1, 5)]

p1 <- p1 + coord_cartesian(ylim = ylim1*4)

plot(p1)
```

As we checked previously, with short flights the delays caused by the carrier are not very big. The results are consistent.

## Late aircraft

```{r plotting San Antonio lateAircraft, message=FALSE}
p2 <- ggplot(ps_by_carrier, aes(Carrier, LateAircraftDelay)) +
  geom_boxplot(outlier.shape=NA) +
  labs(x="Carrier",y="Delay") +
  ggtitle("Delay caused by late aircraft in San Antonio, TX")

ylim1 = boxplot.stats(ps_by_carrier$LateAircraftDelay)$stats[c(1, 5)]

p2 <- p2 + coord_cartesian(ylim = ylim1*1.5)

plot(p2)
```

Again, American Airlines is one of the carriers with most delays caused by a late aircraft. This one can be explained again by the distance of the flights.

## Taxiing

```{r plotting San Antonio TaxiOut, message=FALSE}

p3 <- ggplot(ps_by_carrier, aes(Carrier, TaxiIn+TaxiOut)) +
  geom_boxplot(outlier.shape=NA) +
  labs(x="Carrier",y="Taxi out (minutes") +
  ggtitle("Taxi out in San Antonio, TX")

ylim1 = boxplot.stats(ps_by_carrier$TaxiIn+ps_by_carrier$TaxiOut)$stats[c(1, 5)]

p3 <- p3 + coord_cartesian(ylim = ylim1*1.5)

plot(p3)

```

As we can see, American Airlines in one of the carriers that spends more time taxiing. This is again consistent with previous results.



